---
title: "Visualizing Non-Overlapping Disadvantaged Census Tracts: Bolstering CalEnviroScreen Considerations"
author: "Sofia Ingersoll"
format:
  html:
    code-fold: true
    code-summary: "View the code"
embedded-sources: true
output:
  html_document:
    print_df: paged
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r set-up, message = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)

library(sf)
library(tmap)
library(here)
library(terra)
library(tidyverse)
library(leaflet)
library(tmaptools)
library(osmextract) 
library(RColorBrewer)
```


# Read CalEnviroScreen Data
```{r read_n_wrangle_calenviro}
calenviro <- st_read(here("data","CalEnviroScreen","CES4 Final Shapefile.shp")) %>% 
  janitor::clean_names() %>% 
  st_make_valid()

# housing burden is a good parameter to visualize
# let's get an idea of the bin distribution for plotting
#range(calenviro$hous_burd)

# remove the -999.0, that's an NA
calenviro <- calenviro %>% 
  filter(hous_burd != -999.0)

# check! nice!
#range(calenviro$hous_burd)

calenviro <- st_transform(calenviro,
                          4326) %>% 
  st_make_valid()
```


# Read CDC EJ Data
This raw data only contains CA observations, so no need to filter. However, it lacks geo-info, so we're going to relate the df using `calenviro$tract` and `cdc_ej$fips`

```{r read_n_wrangle_cdc_ej}
cdc_ej <- read_csv(here("data", "CDC_Index", "California_EJ.csv"),
                    show_col_types = FALSE) %>% 
  janitor::clean_names() %>% 
  # match dtype w/ calenviro
  mutate(geoid = as.double(geoid)) %>% 
  rename(tract = geoid) %>% 
  select(tract, everything())

#head(cdc_ej , 3)   
  
# now cdc_svi contains geo-info and all the calenviro info
cdc_ej_geo <- inner_join(cdc_ej, calenviro, by = "tract")
                    
head(cdc_ej_geo , 3)     

# let's make it a spatial object for mapping now
cdc_ej_geo <- st_as_sf(cdc_ej_geo) %>% 
  st_transform(crs = 4326) %>% 
  st_make_valid()
```

```{r cdc_ej_map}
# we'll break housing burden into quantiles
bins <- seq(0, 100, by = 25)

# i want a map will all the data layered for for selection against calenviro
tmap_mode(mode = "view")

overlayed_map <- tm_shape(calenviro) +
  
  tm_basemap(leaflet::providers$OpenStreetMap) +
  
  tm_polygons(col = 'hous_burd',
              palette = 'plasma',
              style = 'fixed',
              breaks = bins,
              title = "Housing Burden") +
  
  tm_shape(cdc_ej_geo) +
  
  # areas with overlap are made to be white,
  # as to "white out" areas not of interest  
  # the only visible regions are where the data do no overlap 
  tm_bubbles(size = .1,
             alpha = 0.5,
             palette = "plasma",
             border.col = 'white'
             ) +
  
   tm_layout(title = "Visualizing the Overlap of CalEnviroScreen and CDC EJ: \\n Highlighting the Housing Burden & Identifying Non-Overlapping Census Tracts",
            title.size = 1.5,
            main.title.position = c("center", "top"),
            legend.title.size = 0.8,
            legend.text.size = 0.7)

tmap_save(overlayed_map, here("Deliverables","interactive_maps","overlayed_maps","overlayed_map_cdc_ej.html"))
```


# Isolate Non-overlapping Tracts

Quality Check:

So the cdc_svi_geo data has 7889 observations that we can map. 
I believe I could bolster this using an additional data set specifically for CA census tract geometries.

When disjoining the data, we expect observations to decrease because our original map had a lot of overlap in the metropolitan areas.
A decrease of 7 observations were removed which is a little confusing comparing to the previous map.
```{r cdc_ej_disjoint_data}
# instead of a sgbp the output
# let's make it a vector,
# this only contains boolean responses for the question we asked above 
no_overlap_vector_cdc_ej <- st_disjoint(cdc_ej_geo, calenviro, sparse = FALSE)[, 1]

#plot(no_overlap_vector_cdc_ej)

#no_overlap_vector_cdc_ej

# subset the calenviro data based on disjoint indices
# now we will only have geoloc data for areas that do not overlap with calenviro
cdc_ej_disjoint <- cdc_ej_geo[no_overlap_vector_cdc_ej, ]

# this is better, we're seeing a drop from 8041 observations to 8034 observations
#cdc_ej_disjoint
```

### Quick Visual Check

Our map here shows a comforting map that has appropriately missing pieces

```{r no_overlap_cdc_ej_plot_check, eval = FALSE}
no_overlap_cdc_ej_map <- ggplot() + 
  
  geom_sf(data = cdc_ej_disjoint) +
  
  theme_minimal() +
  
  labs(title = "Non-Overlapping Disadvantaged Census Tracts",
       subtitle = "CDC EJ vs CalEnviroScreen")

  
no_overlap_cdc_ej_map
```

# Visual Non-overlapping Tracts
```{r cdc_ej_disjoint_map}
# i want a map will all the data layered for for selection against calenviro
tmap_mode(mode = "view")

no_overlap_map <- tm_shape(cdc_ej_disjoint) +
  
  tm_basemap(leaflet::providers$OpenStreetMap) +
  
  tm_polygons(col = 'tot_pop19',
              palette = 'plasma',
              style = 'quantile',
              title = "Population") +
  
  tm_scale_bar() +
  
  tm_layout(title = "Visualizing CDC EJ Non-Overlapping CA Census Tracts with Respect to CalEnviroScreen: \\n Highlighting the Population Density",
            title.size = 5,
            title.position = c("center", "top"),
            legend.title.size = 0.8,
            legend.text.size = 0.7)

tmap_save(no_overlap_map, here("Deliverables","interactive_maps","no_overlap_maps","no_overlap_map_cdc_ej.html"))
```

# Saving Data

Conceptual quams:

I also there to be bigger gaps in the center of the state where there was overlap observed in the first map. However, only 7 observations were removed, so it makes some sense. The initial over-layed map feels deceptive.

```{r save_no_overlap_data}
write_csv(cdc_ej_disjoint, here("Deliverables", "no_overlap_tracts_data", "cdc_ej_no_overlap_tracts.csv"))
```



