---
title: "Visualizing Non-Overlapping Disadvantaged Census Tracts: Bolstering CalEnviroScreen Considerations"
author: "Sofia Ingersoll"
format:
  html:
    code-fold: true
    code-summary: "View the code"
embedded-sources: true
output:
  html_document:
    print_df: paged
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r set-up, message = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)

library(sf)
library(tmap)
library(here)
library(terra)
library(tidyverse)
library(leaflet)
library(tmaptools)
library(osmextract) 
library(RColorBrewer)
```


# Read CalEnviroScreen Data
```{r read_n_wrangle_calenviro}
calenviro <- st_read(here("data","CalEnviroScreen","CES4 Final Shapefile.shp"),
                     quiet = TRUE) %>% 
  janitor::clean_names() %>% 
  st_transform(crs = "EPSG:4326") %>% 
  st_make_valid()

# housing burden is a good parameter to visualize
# let's get an idea of the bin distribution for plotting
#range(calenviro$hous_burd)

# remove the -999.0, that's an NA
calenviro <- calenviro %>% 
  filter(hous_burd != -999.0)

# check! nice!
#range(calenviro$hous_burd)
```

# Read SD CEI Data
```{r read_n_wrangle_sdcei}
sd_cei <- st_read(here("data","sd_climate_equity_index","City_of_San_Diego_CEI_2021_Revision.shp"),
                  quiet = TRUE) %>% 
  janitor::clean_names() %>% 
  st_transform(crs = "EPSG:4326") %>% 
  st_make_valid()

#colnames(sd_cei)

# housing burden is a good parameter to visualize
# let's get an idea of the bin distribution for plotting
#range(calenviro$cei_score)

#unique(sd_cei$cei_score)
```

```{r sdcei_map}
# we'll break housing burden into quantiles
bins <- seq(0, 100, by = 25)

# i want a map will all the data layered for for selection against calenviro
tmap_mode(mode = "view")

overlayed_map <- tm_shape(calenviro) +
  
  tm_basemap(leaflet::providers$OpenStreetMap) +
  
  tm_polygons(col = 'hous_burd',
              palette = 'plasma',
              style = 'fixed',
              breaks = bins,
              title = "Housing Burden") +
  
  tm_shape(sd_cei) +
  
  # areas with overlap are made to be white,
  # as to "white out" areas not of interest  
  # the only visible regions are where the data do no overlap 
  tm_bubbles(size = .5,
             alpha = 0.5,
             palette = "plasma",
             border.col = 'white'
             ) +
  
  tm_scale_bar() +
  
  tm_layout(title = "Visualizing the Overlap of CalEnviroScreen and SD CEI: \\n Highlighting the Housing Burden & Identifying Non-Overlapping Census Tracts",
            title.size = 1.5,
            main.title.position = c("center", "top"),
            legend.title.size = 0.8,
            legend.text.size = 0.7) 

tmap_save(overlayed_map, here("Deliverables","interactive_maps","overlayed_maps","overlayed_map_sd_cei.html"))
```
# Isolate Non-overlapping Tracts

Quality Check:

So the csd_cei data has 297 observations that we can map. 
I believe I could bolster this using an additional data set specifically for CA census tract geometries.

When disjoining the data, we expect observations to decrease because our original map had a lot of overlap in the central valley.
A decrease of 4 observations were removed which is a little confusing comparing to the previous map.
```{r sd_cei_disjoint_data}
# instead of a sgbp the output
# let's make it a vector,
# this only contains boolean responses for the question we asked above 
no_overlap_vector_sd_cei <- st_disjoint(sd_cei, calenviro, sparse = FALSE)[, 1]

#plot(no_overlap_vector_sd_cei)

#no_overlap_vector_sd_cei

# subset the calenviro data based on disjoint indices
# now we will only have geoloc data for areas that do not overlap with calenviro
sd_cei_disjoint <- sd_cei[no_overlap_vector_sd_cei, ]

# this is better, we're seeing a drop from 8041 observations to 8034 observations
#sd_cei_disjoint
```
