---
title: "Visualizing Non-Overlapping Disadvantaged Census Tracts: White House Map vs CalEnviroScreen"
author: "Sofia Ingersoll"
format:
  html:
    code-fold: true
    code-summary: "View the code"
embedded-sources: true
output:
  html_document:
    print_df: paged
    toc: yes
    toc_depth: 4
    toc_float: yes
---

```{r set-up, message = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)

library(sf)
library(tmap)
library(here)
library(terra)
library(tidyverse)
library(leaflet)
library(tmaptools)
library(osmextract) 
library(RColorBrewer)
```


# Read CalEnviroScreen Data
```{r read_n_wrangle_calenviro}
calenviro <- st_read(here("data","CalEnviroScreen","CES4 Final Shapefile.shp")) %>% 
  janitor::clean_names() %>% 
  st_make_valid()

# housing burden is a good parameter to visualize
# let's get an idea of the bin distribution for plotting
#range(calenviro$hous_burd)

# remove the -999.0, that's an NA
calenviro <- calenviro %>% 
  filter(hous_burd != -999.0)

# check! nice!
#range(calenviro$hous_burd)
```

# Read White House Map Data

```{r read_n_wranlge_wh}
white_house <- st_read(here("data", "white_house_map","usa","usa.shp"),
                       quiet = TRUE) %>% 
  janitor::clean_names()

white_house <- white_house %>% 
  filter(sf == "California") %>% 
  rename(tract = geoid10)

# remove empty geometries
white_house <- white_house[!st_is_empty(white_house), ]

white_house <- st_transform(white_house, 
                            crs = crs(calenviro)) %>% 
  st_make_valid()
```

```{r wh_map}
# we'll break housing burden into quantiles
bins <- seq(0, 100, by = 25)

# i want a map will all the data layered for for selection against calenviro
tmap_mode(mode = "view")

overlayed_map <- tm_shape(calenviro) +
  
  tm_basemap(leaflet::providers$OpenStreetMap) +
  
  tm_polygons(col = 'hous_burd',
              palette = 'plasma',
              style = 'fixed',
              breaks = bins,
              title = "Housing Burden") +
  
  tm_shape(white_house) +
  
  # areas with overlap are made to be white,
  # as to "white out" areas not of interest  
  # the only visible regions are where the data do no overlap 
  tm_bubbles(size = .1,
             alpha = 0.5,
             palette = "plasma",
             border.col = 'white'
             ) +
  
  tm_scale_bar() +
  
  
  tm_layout(title = "Visualizing the Overlap of CalEnviroScreen and White House Map: \\n Highlighting the Housing Burden & Identifying Non-Overlapping Census Tracts",
            title.size = 1.5,
            main.title.position = c("center", "top"),
            legend.title.size = 0.8,
            legend.text.size = 0.7)

tmap_save(overlayed_map, here("Deliverables","interactive_maps","overlayed_maps","overlayed_map_whitehouse.html"))
```

# Isolate Non-overlapping Tracts

Quality Check:

So the cdc_svi_geo data has 8036 observations that we can map. 
I believe I could bolster this using an additional data set specifically for CA census tract geometries.

When disjoining the data, we expect observations to decrease because our original map had a lot of overlap in the southern coast.
A decrease of 5 observations were removed which is a little confusing comparing to the previous map.
```{r wh_disjoint_data}
# instead of a sgbp the output
# let's make it a vector,
# this only contains boolean responses for the question we asked above 
no_overlap_vector_wh <- st_disjoint(white_house, calenviro, sparse = FALSE)[, 1]

#plot(no_overlap_vector_wh)

#no_overlap_vector_wh

# subset the calenviro data based on disjoint indices
# now we will only have geoloc data for areas that do not overlap with calenviro
wh_disjoint <- white_house[no_overlap_vector_wh, ]

# this is better, we're seeing a drop from 8041 observations to 8034 observations
wh_disjoint
```

### Quick Visual Check

Our map here shows a comforting map that has appropriately missing pieces

```{r no_overlap_wh_plot_check, eval = FALSE}
no_overlap_wh_map <- ggplot() + 
  
  geom_sf(data = wh_disjoint) +
  
  theme_minimal() +
  
  labs(title = "Non-Overlapping Disadvantaged Census Tracts",
       subtitle = "White House Map vs CalEnviroScreen")

  
no_overlap_wh_map
```
# Visual Non-overlapping Tracts
```{r wh_disjoint_map}
# i want a map will all the data layered for for selection against calenviro
tmap_mode(mode = "view")

no_overlap_map <- tm_shape(wh_disjoint) +
  
  tm_basemap(leaflet::providers$OpenStreetMap) +
  
  tm_polygons(col = 'tpf',
              palette = 'plasma',
              style = 'quantile',
              title = "Population") +
  
  tm_scale_bar() +
  
  tm_layout(title = "Visualizing White House Map Non-Overlapping CA Census Tracts with Respect to CalEnviroScreen: \\n Highlighting the Population Density",
            title.size = 5,
            title.position = c("center", "top"),
            legend.title.size = 0.8,
            legend.text.size = 0.7)

tmap_save(no_overlap_map, here("Deliverables","interactive_maps","no_overlap_maps","no_overlap_map_whitehouse.html"))
```

# Saving Data

Conceptual quams:

I also there to be bigger gaps in the center of the state where there was overlap observed in the first map. However, only 5 observations were removed, so it makes some sense. The initial over-layed map feels deceptive.

```{r save_no_overlap_data}
write_csv(wh_disjoint, here("Deliverables", "no_overlap_tracts_data", "whitehouse_no_overlap_tracts.csv"))
```
