---
title: "Assigning Sunroof Data to the CT Level"
author: "Sofia Ingersoll"
---

```{r set_up, message = FALSE, warning=FALSE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#----             System Set Up         ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Set working directory 
knitr::opts_chunk$set(echo = TRUE)
getwd()

# to ensure observations are reproducible for everyone
set.seed(123)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#----            Load Libraries         ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
library(sf)
library(here)
library(tmap)
library(readxl)
library(leaflet)
library(maptiles)
library(tmaptools)
library(sfheaders)
library(tidyverse)
library(RColorBrewer)
```

Building pv data: join by GEOID + CT.TL for PV data. Ignoring the geospatial data provided by proj. Sunroof. 
Color by yearly_sunlight_kwh_total and gain an intuitive sense of the energy collection. More intense kwh in areas with greater rooftop.
If it’s a 1:1 GEOID match = okay
If not, find data method source or deduce 


```{r read_n_wrangle}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----  Read & Subset P.S.   ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
path_to_sunroof <- '../../../data/Thrust_2.1_Building_Energy_DS/rooftop_solar_potential_data/project-sunroof-census_tract.csv'

sunroof_data <- read_csv(path_to_sunroof,
                         show_col_types = F) %>% 
  janitor::clean_names()

sunroof_ca <- sunroof_data %>% 
  filter(state_name == 'California') %>% 
  rename('GEOID' = 'region_name')

head(sunroof_ca)
```
### Check point

This confirms that we do not have a 1:1 GEOID match. I am still going to assign each observation to the appropriate ct GEOID

Researching quickly before doing any data processing but ideas I have for census tracts with ore than one observation is to average. I feel like that would not be the best option because we are losing data. 

Uh oh, we expect a minimum of at least 8057 census tracts for the state of CA. There is a strong likelihood we are do not have the entire state contained in this data. 

- Contacted the data managers about the missing data between the toggle map website & data download website. Need to find another contact.
- Note that does not account for all missing data: some regions may not meet the qualifications required and defined on the data resource page in our shared drive for Thrust 2.1 of the project. 
```{r ct_check}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----      GEOID Checks     ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# How many ct are repeating?
print(paste("There are",length(sunroof_ca$GEOID), "observations in project sunroof"))
print(paste("There are",length(unique(sunroof_ca$GEOID)), "unique census tracts in project sunroof"))
print(paste("Resulting in",length(sunroof_ca$GEOID)-length(unique(sunroof_ca$GEOID)), "repeating census tracts"))
```

```{r read_tl}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----  Read & Check TL Data ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
census_tract <- st_read(here("data", "Thrust_2.1_Building_Energy_DS","rooftop_solar_potential_data","tl_2023_06_tract", "tl_2023_06_tract.shp"),
                        quiet = T)
#census_tract <- st_read(here("data", "Thrust_2.1_Building_Energy_DS","rooftop_solar_potential_data","tl_2022_06_tract", "tl_2022_06_tract.shp"),
                       # quiet = T)

census_tract

print(paste("There are", length(census_tract$GEOID), "GEOID in the tigerline data"))
```
```{r ct_geometry}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ---- Subset TL GEOID Geoms ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# isolate geoid & geometry for ct
ct <- census_tract %>% 
  select(GEOID, geometry)
```

### Aggregating Geometries

Seeing that some census tracts are repetitive in the Google Sunroof Data, we will aggregate the non-unique observations by GEOID. Our process will include a summation for the count columns respectively and averaging the percentage and median columns respectively. We expect the aggregated output to contain 7504 observations.

```{r sunroof_ct_agg}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----  Standarize P.S. Data ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Convert relevant columns to numeric
sunroof_ca <- sunroof_ca %>%
  mutate(
    GEOID = as.numeric(GEOID),
    yearly_sunlight_kwh_kw_threshold_avg = as.numeric(yearly_sunlight_kwh_kw_threshold_avg),
    count_qualified = as.numeric(count_qualified),
    percent_covered = as.numeric(percent_covered),
    percent_qualified = as.numeric(percent_qualified),
    number_of_panels_n = as.numeric(number_of_panels_n),
    number_of_panels_s = as.numeric(number_of_panels_s),
    number_of_panels_e = as.numeric(number_of_panels_e),
    number_of_panels_w = as.numeric(number_of_panels_w),
    number_of_panels_f = as.numeric(number_of_panels_f),
    number_of_panels_median = as.numeric(number_of_panels_median),
    number_of_panels_total = as.numeric(number_of_panels_total),
    kw_median = as.numeric(kw_median),
    kw_total = as.numeric(kw_total),
    yearly_sunlight_kwh_n = as.numeric(yearly_sunlight_kwh_n),
    yearly_sunlight_kwh_s = as.numeric(yearly_sunlight_kwh_s),
    yearly_sunlight_kwh_e = as.numeric(yearly_sunlight_kwh_e),
    yearly_sunlight_kwh_w = as.numeric(yearly_sunlight_kwh_w),
    yearly_sunlight_kwh_f = as.numeric(yearly_sunlight_kwh_f),
    yearly_sunlight_kwh_median = as.numeric(yearly_sunlight_kwh_median),
    yearly_sunlight_kwh_total = as.numeric(yearly_sunlight_kwh_total),
    carbon_offset_metric_tons = as.numeric(carbon_offset_metric_tons),
    existing_installs_count = as.numeric(existing_installs_count)
  )


#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----  Aggregate P.S. Data  ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Aggregate sunroof_data by GEOID
sunroof_ca_agg <- sunroof_ca %>%
  group_by(GEOID) %>%
  summarize(
    GEOID = first(GEOID),  # Retain first GEOID
    state_name = first(state_name),  # Retain first state_name
    yearly_sunlight_kwh_kw_threshold_avg = mean(yearly_sunlight_kwh_kw_threshold_avg, na.rm = TRUE),  # Mean of threshold avg
    count_qualified = sum(count_qualified, na.rm = TRUE),  # Sum of count_qualified
    percent_covered = mean(percent_covered, na.rm = TRUE),  # Mean of percent_covered
    percent_qualified = mean(percent_qualified, na.rm = TRUE),  # Mean of percent_qualified
    number_of_panels_n = sum(number_of_panels_n, na.rm = TRUE),  # Sum of number_of_panels_n
    number_of_panels_s = sum(number_of_panels_s, na.rm = TRUE),  # Sum of number_of_panels_s
    number_of_panels_e = sum(number_of_panels_e, na.rm = TRUE),  # Sum of number_of_panels_e
    number_of_panels_w = sum(number_of_panels_w, na.rm = TRUE),  # Sum of number_of_panels_w
    number_of_panels_f = sum(number_of_panels_f, na.rm = TRUE),  # Sum of number_of_panels_f
    number_of_panels_median = mean(number_of_panels_median, na.rm = TRUE),  # Mean of number_of_panels_median
    number_of_panels_total = sum(number_of_panels_total, na.rm = TRUE),  # Sum of number_of_panels_total
    kw_median = mean(kw_median, na.rm = TRUE),  # Mean of kw_median
    kw_total = sum(kw_total, na.rm = TRUE),  # Sum of kw_total
    yearly_sunlight_kwh_n = sum(yearly_sunlight_kwh_n, na.rm = TRUE),  # Sum of yearly_sunlight_kwh_n
    yearly_sunlight_kwh_s = sum(yearly_sunlight_kwh_s, na.rm = TRUE),  # Sum of yearly_sunlight_kwh_s
    yearly_sunlight_kwh_e = sum(yearly_sunlight_kwh_e, na.rm = TRUE),  # Sum of yearly_sunlight_kwh_e
    yearly_sunlight_kwh_w = sum(yearly_sunlight_kwh_w, na.rm = TRUE),  # Sum of yearly_sunlight_kwh_w
    yearly_sunlight_kwh_f = sum(yearly_sunlight_kwh_f, na.rm = TRUE),  # Sum of yearly_sunlight_kwh_f
    yearly_sunlight_kwh_median = mean(yearly_sunlight_kwh_median, na.rm = TRUE),  # Mean of yearly_sunlight_kwh_median
    yearly_sunlight_kwh_total = sum(yearly_sunlight_kwh_total, na.rm = TRUE),  # Sum of yearly_sunlight_kwh_total
    carbon_offset_metric_tons = sum(carbon_offset_metric_tons, na.rm = TRUE),  # Sum of carbon_offset_metric_tons
    existing_installs_count = sum(existing_installs_count, na.rm = TRUE)  # Sum of existing_installs_count
  )

# View the aggregated data
sunroof_ca_agg
```
### Join GEOID of Project Sunroof & TIGER/Line

```{r join_sunroof_tl}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----     Join P.S. + TL    ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
ct_pv <- ct %>%
  mutate(GEOID = as.numeric(GEOID)) %>% 
  left_join(sunroof_ca_agg, by = "GEOID")

ct_pv

# save as spatial object
ct_pv <- st_as_sf(ct_pv) %>% 
  st_transform(4269) %>% 
  st_make_valid()

print(st_geometry(ct_pv))
```

```{r summary_stats_pstl}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----  Join Summary Stats   ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# project sunroof tigerline summary stats
summary(ct_pv$yearly_sunlight_kwh_total)
quantile(ct_pv$yearly_sunlight_kwh_total, na.rm =T)
```

### Plot Sunroof + TIGER/Line
```{r map_yearly_sunlight_kwh_tot_grey_na}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----  Map Yearly Sun/kWH   ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Calculate 4 quantiles
quantile_breaks <- quantile(ct_pv$yearly_sunlight_kwh_total, probs = seq(0, 1, 0.25), na.rm = TRUE)

# Create custom labels with kWh ranges
quantile_labels <- paste0(
  format(round(quantile_breaks[-length(quantile_breaks)] / 1e6, 2), big.mark = ",", scientific = FALSE), 
  "M - ",
  format(round(quantile_breaks[-1] / 1e6, 2), big.mark = ",", scientific = FALSE), 
  "M kWh"
)

# Create 4 quantile groups with the custom labels
ct_pv_quant <- ct_pv %>%
  mutate(
    sunlight_quantile = cut(
      yearly_sunlight_kwh_total,
      breaks = quantile_breaks,
      include.lowest = TRUE,
      labels = quantile_labels
    )
  )

# Plot using the quantiles
ggplot() +
  geom_sf(data = ct_pv_quant, aes(fill = sunlight_quantile)) +
  scale_fill_brewer(
    # Sequential color scale (Blues)
    palette = "Blues",  
    # Color for NA values
    na.value = "grey50",  
    # Use the custom labels with kWh ranges
    labels = quantile_labels 
  ) +
  labs(
    title = "Yearly Sunlight Energy Potential (kWh) by Quantiles",
    fill = "Yearly Sunlight (kWh)"
  ) +
  theme_minimal() +
  theme(
    # Larger, readable text
    legend.text = element_text(size = 10), 
    # Increase height of legend items
    legend.key.height = unit(1.5, "cm"),          
    # Increase width of legend keys
    legend.key.width = unit(2, "cm"),        
    # More vertical space between legend items
    legend.spacing.y = unit(0.7, "cm"),                   
    # Increase legend title size for readability
    legend.title = element_text(size = 11),  
    # Position the legend to the right
    legend.position = "right"                             
  )
```


```{r existing_ct_no_na_map}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----      Map Without NA   ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Remove missing data
ct_pv_filtered <- ct_pv %>% filter(!is.na(yearly_sunlight_kwh_total))

# Calculate 4 quantiles
quantile_breaks <- quantile(ct_pv_filtered$yearly_sunlight_kwh_total, probs = seq(0, 1, 0.25), na.rm = TRUE)

# Create custom labels with kWh ranges
quantile_labels <- paste0(
  format(round(quantile_breaks[-length(quantile_breaks)] / 1e6, 2), big.mark = ",", scientific = FALSE), 
  "M - ",
  format(round(quantile_breaks[-1] / 1e6, 2), big.mark = ",", scientific = FALSE), 
  "M kWh"
)

# Create 4 quantile groups with the custom labels
ct_pv_filtered <- ct_pv_filtered %>%
  mutate(
    sunlight_quantile = cut(
      yearly_sunlight_kwh_total,
      breaks = quantile_breaks,
      include.lowest = TRUE,
      labels = quantile_labels
    )
  )

# Plot using the quantiles
ggplot() +
  geom_sf(data = ct_pv_filtered, aes(fill = sunlight_quantile)) +
  scale_fill_brewer(
    palette = "Oranges",  
    na.value = "grey50",
    labels = quantile_labels  
  ) +
  labs(
    title = "Yearly Sunlight Energy Potential (kWh) by Quantiles",
    fill = "Yearly Sunlight (kWh)"
  ) +
  theme_minimal() +
  theme(
    legend.text = element_text(size = 10),           
    legend.key.height = unit(1.5, "cm"),         
    legend.key.width = unit(2, "cm"),          
    legend.spacing.y = unit(0.7, "cm"),                
    legend.title = element_text(size = 11),             
    legend.position = "right"                             
  )
```

### Save Data

We can see using plotting above and Esri Mapping products like QGIS to visualize the sparseness of the data.
```{r eval = FALSE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----       Save Data       ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Create 4 quantile groups with the custom labels
ct_pv_filtered_2 <- ct_pv %>% filter(!is.na(yearly_sunlight_kwh_total))

# Convert to sf object if not already
ct_pv_sf <- st_as_sf(ct_pv_filtered_2, coords = "geometry", remove = FALSE)

# Save as GeoJSON
file_path <- here("data", "Thrust_2.1_Building_Energy_DS","rooftop_solar_potential_data", "census_tract_pv","census_tract_pv.geojson")

st_write(ct_pv_sf, file_path, driver = "GeoJSON")

# Save as Shapefile
file_path2 <- here("data", "Thrust_2.1_Building_Energy_DS","rooftop_solar_potential_data", "census_tract_pv","census_tract_pv.shp")

st_write(ct_pv_sf, file_path2, driver = "ESRI Shapefile")
```

### What's going on with the Missing CT?

Let's inspect the problem areas, why are they empty?
Are the GEOID not matching?

First we'll inspect how the discrepancies between Project Sunroof and Census TIGER/Line GEOID and get a sense of how many are missing.
- It's important to note that TIGER/Line includes additional boundaries for rivers and special geographic features. This is why there are more than 8,057 observations in the TIGER/Line object. 
- We found 1,042 observations in the TIGER/Line are not accounted for in the Project Sunroof data.
- When comparing the GEOIDs that did not match, 0 observations were missing from the Project Sunroof data. This confirms our suspicion that these additional observations are geographical features in the TIGER/Line.

To get a more accurate understanding of how many census tracts are properly represented, we're going to compare the Google Project Sunroof to a list of the 8,057 CA census tract GEOIDs. 
- Upon inspection, only one GEOID from the Project Sunroof was not accounted for in the CA census tract list.


Now, let's join census tract TIGER/Line with the CA tract list to see what is matching vs missing. 
- The number of GEOID in the TIGER/Line not present in the CA Census Tract list is 2248 and we expected there to be 1072 missing GEOID. Therefore, 1176 census tract GEOID are missing from the TIGER/Line data.
- Further validation proves the number of GEOID in the CA Census Tract list not present in the TIGER/Line is 1176.

Comparing the CA Tract List data to Project Sunroof GEOID revealed there are 7504 unique GEOIDs in the Project Sunroof Data.
- When combining with the CA Census Tract List, 7503 GEOIDs are accounted for.
- When combining this data with the TIGER/Line geometries a simple feature collection with 7503 features and 24 fields (with 1041 geometries empty) are produced. Providing geometry for only 6462 census tracts.

```{r sunroof_ct_vs_tiger_ct}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ---- Identify Missing GEOID ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# isolate unique geoid
sunroof_geoids <- unique(sunroof_ca$GEOID) %>% 
  as.data.frame() 

colnames(sunroof_geoids) <- "GEOID"

head(sunroof_geoids, 3)

ct_geoids <- unique(ct_pv$GEOID) %>% 
  as.data.frame() 

colnames(ct_geoids) <- "GEOID"

head(ct_geoids, 3)

# what geoids are different?
geoids_in_sunroof_not_ct <- setdiff(sunroof_geoids, ct_geoids)

geoids_in_sunroof_not_ct


#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----  P.S. Missing GEOID   ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# let's select these geoids from sunroof_ca that are not in ct tigerline
# 0 bs, all census tracts are accounted for from the project sunroof data
sunroof_ca_missing_in_ct <- sunroof_ca %>%
  filter(GEOID %in% geoids_in_sunroof_not_ct)

# 0 observations confirms our suspicion that these additional observations are geographical features in the TIGER/Line
print(sunroof_ca_missing_in_ct)
```

```{r census_tract_list}
#~~~~~~~~~~~~~~~~~~~~~~~~
# ----  tract data   ----
#~~~~~~~~~~~~~~~~~~~~~~~~
# path to your Excel file
tract_path <- here("data", "Thrust_1.1_DAC_Survey","Tract-level-CA-HTC-Index-for-public-website-download-20190304.xlsx")

# names of all sheets in the Excel file
sheet_names <- readxl::excel_sheets(tract_path)

# read the fourth sheet directly by its index
tract_list <- readxl::read_xlsx(tract_path, sheet = sheet_names[1]) %>% 
  janitor::clean_names() %>%
  as.data.frame() %>% 
  select(geoid) %>%   
  mutate(GEOID = geoid) %>% 
  distinct(GEOID) %>% 
  mutate(GEOID = as.numeric(GEOID))

# 8057 obs
tract_list
```

```{r sunroof_ct_vs_tract_list}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----   P.S. Missing GEOID  ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# what geoids are different? In P.S. not, in tractlist
geoids_in_sunroof_not_tract_list <- setdiff(sunroof_geoids, tract_list)

geoids_in_sunroof_not_tract_list

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ---- Tractlist Missing GEOID ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# what's missing from the complete tract list?
tract_list_geoids_not_in_sunroof <- setdiff(tract_list, sunroof_geoids)

# As expected, 554 CA census tracts are not accounted for in the project sunroof data 
tract_list_geoids_not_in_sunroof
```

```{r tigerline_vs_ca_tract_list}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----    T.L. Missing GEOID   ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# what geoids are different? In TL not, in tractlist
# find the GEOID differences between ct_geoids and tract_list
missing_geoids <- setdiff(ct_geoids$GEOID, tract_list$GEOID) %>% 
  as.data.frame()

colnames(missing_geoids) <- "GEOID"

# check the results for geoid in TL not tractlist
print(paste("The number of GEOID in the TIGER/Line not present in the CA Census Tract list is", 
            # compare length of missing ct
            length(missing_geoids$GEOID), 
            "and we expected there to be", 
            # expectation comparing tract list (8057) to T.L (9129)
            (length(unique(ct_geoids$GEOID)) - length(unique(tract_list$GEOID))), 
            "missing GEOID. Therefore,",
            # comparing the number of missing ct with the expected size of the output
            (nrow(missing_geoids) - (length(unique(ct_geoids$GEOID)) - length(unique(tract_list$GEOID)))),
            "census tract GEOID are missing from the TIGER/Line data."))

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ---- Tractlist Missing GEOID ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# what geoids are different? In tract list not, in TIGER/Line
# find the GEOID differences between ct_geoids and tract_list
comparing_geoids <- setdiff(tract_list$GEOID, ct_geoids$GEOID)%>% 
  as.data.frame()

colnames(comparing_geoids) <- "GEOID"

# check the results
print(paste("Further validation proves the number of GEOID in the CA Census Tract list not present in the TIGER/Line is", 
            length(comparing_geoids$GEOID),"."))
```

```{r join_tractlist_ct_pv_filter_na}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# ----  Join Census Tracts  ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# let's join the data and visualize it 
# this contains values 
tract_list_joined <-  left_join(tract_list, ct_pv, by = "GEOID") 

tract_list_joined

tract_list_filtered <- tract_list_joined %>%
  filter(!is.na(yearly_sunlight_kwh_total))

tract_list_filtered 

#plot(tract_list_joined$GEOID)
# are these identical with the sunroof_ca_missing_in_ct?
#missing_ct <- setdiff(sunroof_ca_missing_in_ct$GEOID, missing_geoids$GEOID)
```

When combining the CA Census Tract List with the Project Sunroof Data, 7503/8057 

When combining CA Census Tract List + Solar Data with the TIGER/Line data, there are 1041 empty geometries provided.

```{r combining_all_data}
# first combine the census tract information with the aggregated sunroof data
tract_list_solar <-  left_join(tract_list, sunroof_ca_agg, by = "GEOID") 

# next remove all the observations that did not match
tract_solar_filtered <- tract_list_solar %>%
  filter(!is.na(yearly_sunlight_kwh_total))

# quick check
print(paste("There are",
      nrow(sunroof_ca_agg),
      "unique GEOIDs in the Project Sunroof Data. When combining with the CA Census Tract List,",
      nrow(tract_solar_filtered),
      "GEOIDs are accounted for. Now, when combining this data with the TIGER/Line geometries"))

tract_solar_test <- ct %>% 
  mutate(GEOID = as.numeric(GEOID)) %>% 
  right_join(tract_solar_filtered, by = "GEOID") 


tract_solar_test
```

### New Census Boundary data

Using White House Map Data
```{r}
ct_fresh <- st_read(here("data", "Thrust_1.1_DAC_Survey","white_house_map","usa", "usa.shp"),
                        quiet = T) 

ct_fresh <- ct_fresh %>% 
  filter(SF == "California") %>% 
  mutate(GEOID = as.numeric(GEOID10))

ct_fresh

# what geoids are different? In TL not, in tractlist
# find the GEOID differences between ct_geoids and tract_list
missing_geoid <- setdiff(ct_fresh$GEOID, tract_list$GEOID) %>% 
  as.data.frame()

colnames(missing_geoid) <- "GEOID"

# check the results
print(paste("The number of GEOID in the White House Map data not present in the CA Census Tract list is", 
            length(missing_geoid$GEOID), 
            "and we expect there to be", 
            (length(unique(ct_fresh$GEOID)) - length(unique(tract_list$GEOID))), 
            "missing GEOID."))

# quick comparison of whitehouse data and solar data geoid
# fabulous, it contains exactly what we need!
extra_geoid <- setdiff(ct_fresh$GEOID, sunroof_ca_agg$GEOID) %>% 
  as.data.frame()

colnames(extra_geoid) <- "GEOID"

extra_geoid
```

### White House Map Data Joined with Solar PV Data for CA!

They are a perfect match! Let's put it all together, map and save!
```{r join_whmap_solar_pv}
# inner join
solar_census <- inner_join(sunroof_ca_agg, ct_fresh, by = "GEOID")

solar_census_sf <- st_as_sf(solar_census) %>% 
  st_transform(4326) %>% 
  st_make_valid()

solar_census_sf

#plot(solar_census_sf)


# save as shapefile
file_path <- here("data", "Thrust_2.1_Building_Energy_DS","rooftop_solar_potential_data", "census_tract_pv_whitehouse_data","census_tract_pv_whitehouse_data.gpkg")

st_write(solar_census_sf, file_path, driver = "GPKG")
```


### SCRATCH WORK BELOW
```{r missing_ct}
missing_geoids <- ct_pv %>%
  filter(is.na(yearly_sunlight_kwh_total)) %>%
  select(GEOID)

# View the unmatched GEOIDs
print(missing_geoids)

# Identify duplicate GEOIDs
duplicates <- sunroof_ca %>%
  group_by(GEOID) %>%
  filter(n() > 1)

duplicates
```


```{r join}
# let's inspect the problem areas, why are they empty?
# are the geoid not matching?
empty_obs <- ct_pv %>% 
  filter(st_is_empty(ct_pv) == TRUE)

empty_obs <- empty_obs %>% 
  arrange(desc(GEOID))

empty_obs
```


```{r}
ggplot() +
  geom_sf(data = ct_pv, aes(fill = yearly_sunlight_kwh_total)) +
  scale_fill_viridis_c(option = "plasma", na.value = "grey50") +
  labs(title = "Yearly Sunlight Energy Potential",
       fill = "kWh/year") +
  theme_minimal()
```

```{r join}
file_path <- here("data", "Thrust_2.1_Building_Energy_DS","rooftop_solar_potential_data", "census_tract_pv","census_tract_pv.gpkg")
# save our data for mapping later
st_write(ct_pv, file_path)
```
```{r}
print(st_geometry(ct_pv))
```


```{r}
plot(ct_pv$yearly_sunlight_kwh_total)

plot(ct_pv$geometry)
```

