---
title: "LEAD Data Inspection"
author: "Sofia Ingersoll"
---

```{r set_up, message = FALSE, warning=FALSE}
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#----             System Set Up         ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Set working directory 
knitr::opts_chunk$set(echo = TRUE)
getwd()

# to ensure observations are reproducible for everyone
set.seed(123)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
#----            Load Libraries         ----
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
library(sf)
library(here)
library(terra)
library(tmap)
library(leaflet)
library(maptiles)
library(tmaptools)
library(sfheaders)
# for origin-destination data methods
library(sfnetworks)  
library(tidyverse)
library(RColorBrewer)
```

This data provides aggregated summaries for census tract homes and provides the demographic information associated. It does not include the geolocations of these homes and does not include building type or ownership status. 
```{r}
# Re-read the data, skipping the first 8 rows
lead_data <- read_csv(
  here("data", "Thrust_2.1_Building_Energy_DS", "building_type_data", "LEAD Tool Data Census Tracts (Aug 22, 2024 9_38am).csv"),
  skip = 8,  # Skip the first 8 rows
  quote = "\"",  # Specify quote character if needed
  na = c("", "NA"),  # Handle missing values
  trim_ws = TRUE  # Trim whitespace from characters
)

lead_data$`Geography ID` <- str_remove(lead_data$`Geography ID`, "^0+")

lead_data <- lead_data %>% 
  rename(GEOID = `Geography ID`) %>% 
  mutate(GEOID = as.numeric(GEOID))

lead_data
```

#### Household Characteristics

Households can be broken down by: 

    Area Median Income (AMI)1
    State Median Income (SMI)1
    Federal Poverty Level (FPL)2
    Race, Household Income, Education, & Demographic

1. A MI and SMI categories: 0-30%, 30-50%, 50-80%, 80-100%, 100+% 
2. FPL categories: 0-100%, 100-138%, 138%-200%, 200-400%, 400%+

#### Housing Unit Characteristics

Users can analyze energy expenditures by: 

    Occupant type (owner, renter) 
    Building age
    Building type
    Number of units in building
    Primary heating fuel type
    
### FPL Data overview

This does not contain exact location, only census tract level data

```{r read_fpl}
lead_fpl_data <-  read_csv(here("data", "Thrust_2.1_Building_Energy_DS", "building_type_data", "CA-2018-LEAD-data", "CA FPL Census Tracts 2018.csv"))

lead_fpl_data <- lead_fpl_data %>% 
  rename(GEOID = FIP) %>% 
  mutate(GEOID = as.numeric(GEOID))

head(lead_fpl_data)
```
### SMI Data overview

This does not contain exact location, only census tract level data

```{r read_smi}
lead_smi_data <-  read_csv(here("data", "Thrust_2.1_Building_Energy_DS", "building_type_data", "CA-2018-LEAD-data", "CA SMI Census Tracts 2018.csv"))

lead_smi_data <- lead_smi_data %>% 
  rename(GEOID = FIP) %>% 
  mutate(GEOID = as.numeric(GEOID))

head(lead_smi_data)
```
### AMI Data overview

This does not contain exact location, only census tract level data

```{r read_smi}
lead_ami_data <-  read_csv(here("data", "Thrust_2.1_Building_Energy_DS", "building_type_data", "CA-2018-LEAD-data", "CA AMI Census Tracts 2018.csv"))

lead_ami_data <- lead_ami_data %>% 
  rename(GEOID = FIP) %>% 
  mutate(GEOID = as.numeric(GEOID))

head(lead_ami_data)
```

### Combining Query and FPL data with census tract .shp data to map
```{r joining}
combined_lead <- full_join(lead_data, lead_fpl_data, by = "GEOID")

head(combined_lead)
```

```{r read_spatial}
#~~~~~~~~~~~~~~~~~~~~~~~~
# ----  tract data   ----
#~~~~~~~~~~~~~~~~~~~~~~~~
# path to your Excel file
tract_path <- here("data", "Thrust_1.1_DAC_Survey","tl_2021_06_tract", "tl_2021_06_tract.shp")

# names of all sheets in the Excel file
ct_sf <- st_read(tract_path) %>% 
  st_transform(4326) %>% 
  st_make_valid()


ct_sf$GEOID <- str_remove(ct_sf$GEOID, "^0+")

ct_sf
```

```{r join_spatial}
lead_sf <- full_join(ct_sf, combined_lead, by = "GEOID")

```

```{r map}

```

